\chapter{Pelajarilah Seninya} %81

\bahasa
17 Februari 1979 pagi di Aula Buddha

\english
17 February 1979 am in Buddha Hall

\bahasa
Pertanyaan pertama:

\english
The first question:

\bahasa
OSHO TERKASIH,
PIKIRANKU MENGALAMI KESULITAN UNTUK MEMAHAMI ADANYA TANPA-AWAL. MAUKAH ENGKAU BERBICARA TENTANG ITU?

\english
BELOVED OSHO,
MY MIND HAS DIFFICULTY UNDERSTANDING THE EXISTENCE OF BEGINNINGLESSNESS. WOULD YOU PLEASE TALK ABOUT IT?

\bahasa
Prem Viraj, kenyataan tidak dapat didefinisikan. Kenyataan hanya ada; tidak ada cara untuk mengatakan apa itu. Kenyataan bukan "apa," bahkan bukan "itu", kenyataan adalah ini . Kenyataan adalah kebeginiannya: engkau bisa mengalaminya, tapi itu tidak bisa dijelaskan.

\english
Prem Viraj, reality is indefinable. Reality simply is; there is no way to say what it is. It is not a "what," it is not even a "that," it is this. It is this-ness: you can experience it, but it cannot be explained.

\bahasa
Dan kenyataan tanpa awal dan tanpa akhir. Pikiran memiliki awal dan akhir, maka pikiran dan kenyataan tidak dapat bertemu. Pikiran tidak bisa memahami yang abadi. Pikiran ada pada satuan waktu, sebenarnya pikiran adalah waktu; itu ada di masa lalu dan masa depan. Dan ingat, waktu hanya terdiri dari dua masa, masa lalu dan masa depan. Saat ini bukan bagian dari waktu, saat ini adalah bagian dari keabadian.

\english
And reality is beginningless and endless. The mind has a beginning and an end, hence the mind and reality cannot meet. The mind cannot comprehend the eternal. The mind exists in time, in fact the mind is time; it exists in the past and the future. And remember, time consists of only two tenses, the past and the future. The present is not part of time, the present is part of eternity.

\bahasa
Oleh karena itu pikiran tidak pernah ditemukan di masa sekarang. Itu selalu bergerak, baik terhadap masa lalu maupun masa depan. Pikiran bergerak ke dalam apa yang tidak, atau ke dalam apa yang belum. Seluruh keterampilannya terdiri dari bermimpi. Pikiran berakar pada yang tidak nyata, oleh karena itu pikiran tidak dapat memahami semesta. Analoginya seperti kegelapan. Bagaimana kegelapan bisa memahami cahaya? Bagaimana kematian bisa memahami hidup? Jika kematian bisa memahami hidup maka kematian pasti kehidupan. Jika kegelapan memahami cahaya maka kegelapan pasti cahaya.

\english
Hence the mind is never found in the present. It is always wavering, either towards the past or towards the future. It moves into that which is not, or into that which is not yet. Its whole skill consists in dreaming. It is rooted in the nonexistential, hence it cannot understand existence itself. It is like darkness. How can darkness comprehend light? How can death comprehend life? If death can comprehend life then death will have to be life. If darkness has to comprehend light then darkness will have to be light.

\bahasa
Demikian juga halnya dengan pikiran. Jika pikiran ingin memahami kenyataan, ia harus keluar dari masa lalu dan masa depan. Tapi keluar dari masa lalu dan masa depan, itu tidak lagi menjadi pikiran sama sekali. Oleh karena itu desakan semua master besar di dunia bahwa pintu menuju kenyataan adalah tanpa-pikiran.

\english
And so is the case with the mind. If the mind wants to comprehend reality, it will have to come out of the past and the future. But coming out of the past and the future, it is no longer the mind at all. Hence the insistence of all the great masters of the world that the door to reality is no-mind.

\bahasa
Keluar dari pikiran, dan engkau akan tahu apa itu: yang tak berawal, yang tak berakhir. Tetap terbatas pada pikiran, dan engkau akan bingung; kenyataan tetap tak terbayangkan

\english
Slip out of the mind, and you will know what it is: the beginningless, the endless. Remain confined to the mind, and you will be puzzled; reality remains inconceivable.

\bahasa
Aku tidak bisa menjelaskannya kepadamu, karena dalam menjelaskan aku harus menggunakan pikiran. Mencoba memahaminya, engaku harus menggunakan pikiran. Aku bisa hening denganmu, dan jika engkau juga bisa hening bersamaku maka ada pengertian. Pengertian bukan pikiran. Lalu ada kecerdasan yang besar, wawasan yang besar. Tiba-tiba engaku tahu, dan engkau tahu dari pusat keberadaanmu yang sama sekali berbeda; engkau tahu dari hati. Pengetahuanmu memiliki kualitas cinta, bukan pengetahuan. Pengetahuanmu memiliki kualitas transendensi. Itu sudah bukan lagi pengetahuan ilmiah, bisa disusutkan menjadi konsep. Itu adalah visi puitis, itu adalah pengalaman mistis.

\english
I cannot explain it to you, because in explaining I will have to use the mind. Trying to understand it, you will have to use the mind. I can be silent with you, and if you can also be silent with me then there is understanding. Understanding is not of the mind. Then there is great intelligence, great insight. Suddenly you know, and you know from a totally different center of your being; you know from the heart. Your knowing has the quality of love, not of knowledge. Your knowing has the quality of transcendence. It is no longer scientific knowledge, reducible to concepts. It is poetic vision, it is mystic experience.

\bahasa
Viraj, jika engkau benar-benar ingin mengerti, engkau harus kehilangan pikiran itu. Engkau harus membayar harga kehilangan pikiran itu. Tapi jika engkau bersikeras, "Aku harus mengerti melalui pikiran", maka hanya satu hal yang mungkin terjadi. Pikiran akan meyakinkanmu, perlahan-lahan, bahwa tidak ada yang tidak berawal, tidak ada yang tiada akhir, tidak ada yang tidak dapat didefinisikan, tidak ada yang tidak dapat diketahui.

\english
Viraj, if you really want to understand, you will have to lose the mind. You will have to pay the price of losing the mind. But if you insist, "I have to understand through the mind," then only one thing is possible. The mind will convince you, slowly slowly, that there is nothing which is beginningless, nothing which is endless, nothing which is indefinable, nothing which is unknowable.

\bahasa
Pikiran akan menyusutkan pengalamanmu menjadi sesuatu yang dapat diukur, dapat dimengerti, dapat diketahui -- dan yang dapat diketahui adalah normal, biasa-biasa saja. Yang tidak diketahui adalah suci. Dan hanya dengan yang tidak diketahui hidup menjadi sebuah ucapan syukur, hanya dengan yang tidak diketahui, engkau bergetar dengan keajaiban hidup dan semesta. Tiba-tiba sebuah lagu terlahir di hatimu -- sebuah lagu yang tidak bisa dibatasi, sebuah lagu yang mulai meluap, sebuah lagu yang mulai menjangkau orang lain. Tarian terlahir di dalam dirimu -- tarian yang harus dibagikan, tarian yang tidak bisa membuatmu kikir, tarian yang membuatmu murah hati. Cinta lahir di dalam dirimu -- cinta yang begitu tak terbatas sehingga bisa memenuhi seluruh semesta yang tak terbatas. Itu adalah pemahaman yang nyata. Tapi itu terjadi hanya saat pikiran dijatuhkan.

\english
The mind will reduce your experience to the measurable, the fathomable, the knowable -- and the knowable is ordinary, mundane. The unknowable is sacred. And only with the unknowable does life become a benediction, only with the unknowable are you thrilled with the wonder of life and existence. Suddenly a song is born in your heart -- a song that cannot be contained, a song that starts overflowing, a song that starts reaching others. A dance is born in you -- a dance that has to be shared, a dance about which you cannot be miserly, a dance that makes you generous. A love is born in you -- a love that is so infinite that it can fill the whole infinite existence. That is real understanding. But it happens only when the mind is dropped.

\bahasa
Jangan mencoba melakukan hal yang tidak mungkin. Mencoba memahami kenyataan melalui pikiran ibarat mengangkat dirimu keatas dengan menggunakan tali sepatamu sendiri. Mungkin engkau bisa melompat sedikit, tapi melompat tidak akan membantu; engaku akan kembali ke tanah lagi dan lagi, dan itu akan sangat melelahkan. Hanya dengan menarik tali sepatumu sendiri, engkau tidak bisa terbang ke angkasa; itu tidak akan memberimu sayap.

\english
Don't try to do the impossible. Trying to understand reality through the mind is like pulling yourself up by your own shoestrings. Maybe you can hop a little bit, but that hopping is not going to help; you will be back on the ground again and again, and it will be very tiring. Just by pulling your own shoestrings you cannot fly into the sky; that is not going to give you wings.

\bahasa
Perlahan-lahan, pelajari seni untuk terhubung dengan realitas tanpa pikiran mencampuri. Terkadang saat matahari terbenam, duduk saja melihat matahari, tidak berpikir tentang itu -- menyaksikan, tidak mengevaluasi, bahkan tidak mengatakan, "Betapa indahnya itu!" Saat engkau mengatakan sesuatu, pikiran telah masuk.

\english
Slowly slowly, learn the art of contacting reality without the mind interfering. Sometimes when the sun is setting, just sit there looking at the sun, not thinking about it -- watching, not evaluating, not even saying, "How beautiful it is!" The moment you say something, the mind has come in.

\bahasa
Pikiran terdiri dari bahasa. Jangan gunakan bahasa. Tidak bisakah engkau melihat matahari terbenam dan keindahannya? Tidak bisakah engkau diliput oleh keindahannya? Tidak bisakah engkau dimiliki oleh kemegahannya? Apa gunanya membawa bahasa? Tidak ada yang memintamu untuk mengatakan apa pun. Matahari tidak mengerti bahasamu, awan yang begitu indah dan bercahaya di bawah sinar matahari tidak dapat memahami bahasamu. Mengapa membawanya masuk? Sisihkan kata-kata; bersentuhan langsung, jadilah bergetar. Jika air mata muncul di matamu, bagus. Jika engkau mulai menari, bagus. Atau jika engkau tetap tidak bergerak, terpaku ke keindahan matahari, mabuk, engkau akan mendapatkan sedikit pengalaman -- sedikit pengalaman yang berjalan sangat jauh, sedikit sekilas tanpa pikiran.

\english
The mind consists of language. Don't use language. Can't you just see the sunset and its beauty? Can't you be overwhelmed by its beauty? Can't you be possessed by its grandeur? What is the point of bringing language in? Nobody is asking you to say anything. The sun does not understand your language, the clouds that have become so beautiful and luminous in the setting sun are unable to understand your language. Why bring it in? Put it aside; be in direct contact, be thrilled. If tears come to your eyes, good. If you start dancing, good. Or if you simply remain unmoving, stoned on the beauty of the sun, intoxicated, you will have gained a little experience -- a little experience that goes very far, a little glimpse of no-mind.

\bahasa
Dan ada seribu satu situasi setiap hari. Memegang tangan wanitamu atau priamu, tidak perlu untuk berbicara. Orang terus berbicara - yakkety-yakketyyakkety. Dan alasan mengapa mereka berbicara adalah mereka takut untuk menjadi hening, mereka takut melihat kebenaran, mereka takut melihat kekosongan mereka, mereka takut untuk menyingkapkan diri mereka sendiri, mereka takut untuk melihat jauh ke dalam yang lain. Berbicara terus menerus membuat mereka tetap di permukaan, terisi, sibuk.

\english
And there are a thousand and one situations every day. Holding the hand of your woman or your man, there is no need to talk. People are continuously talking -- yakkety-yakketyyakkety. And the reason why they are talking is that they are afraid to be silent, they are afraid to see the truth, they are afraid to see their utter emptiness, they are afraid to expose themselves, they are afraid to look deep into the other. Continuous talking keeps them on the surface, occupied, engaged.

\bahasa
Memegang tangan wanita atau pria mu, mengapa tidak duduk diam? Mengapa tidak menutup mata dan merasakan? Rasakan kehadiran yang lain, masuklah ke kehadiran yang lain, biarkan kehadiran orang lain masuk ke dalam dirimu; bergetar bersama, bergoyang bersama; Jika tiba-tiba ada energi yang besar yang memilikimu, berdansa bersama -- dan engkau akan mencapai puncak sukacita orgasme seperti yang belum pernah engkau ketahui sebelumnya. Puncak orgasme itu tidak ada hubungannya dengan seks, sebenarnya itu banyak berhubungan dengan keheningan.

\english
Holding the hand of your woman or man, why not sit silently? Why not close your eyes and feel? Feel the presence of the other, enter into the presence of the other, let the other's presence enter into you; vibrate together, sway together; if suddenly a great energy possesses you, dance together -- and you will reach to such orgasmic peaks of joy as you have never known before. Those orgasmic peaks have nothing to do with sex, in fact they have much to do with silence.

\bahasa
Dan jika engkau juga bisa menjadi meditativ dalam kehidupan seksmu, jika engkau bisa hening ketika bercinta, dalam semacam tarian, engkau akan terkejut. Engkau memiliki proses bawaan untuk membawamu ke pantai terjauh.

\english
And if you can also manage to become meditative in your sex life, if you can be silent while making love, in a kind of dance, you will be surprised. You have a built-in process to take you to the farthest shore.

\bahasa
Orang-orang bercinta dengan cara yang sangat buruk sehingga jika anak-anak terkadang melihat orang tua mereka bercinta, mereka mengira mereka sedang bergumul, berkelahi - bahwa Ayah akan membunuh Ibu! Mengerang, bernafas dengan cara yang buruk, kekerasan, gerakan mereka tidak memiliki keanggunan. Itu bukan sebuah tarian; tentu itu bukan sebuah tarian.

\english
People make love in such an ugly way that if children sometimes see their parents making love, they think they are wrestling, fighting -- that Daddy is going to kill Mum! Groaning, breathing in an ugly way, violent, their movements have no elegance. It is not a dance; certainly it is not a dance.

\bahasa
Dan kecuali jika seks menjadi tarian itu akan tetap sangat fisiologis; itu tidak akan memiliki spiritualitas di dalamnya. Tapi itu tidak mustahil. Kecuali seluruh hidupmu penuh dengan saat-saat yang datang ketika pikiran berhenti, kehidupan cintamu tidak dapat beralih ke keheningan.

\english
And unless it becomes a dance it will remain very very physiological; it won't have any spirituality in it. But it is impossible. Unless your whole life is saturated with those moments that come when the mind ceases, your love life cannot move into silence.

\bahasa
Malam penuh dengan bintang. Berbaringlah di bumi, menghilang ke bumi. Kita datang dari bumi, suatu hari kita akan kembali ke bumi untuk beristirahat selamanya. Pada malam hari terkadang, berbaring di halaman, menghilang ke bumi. Lihatlah bintang-bintang -- lihat saja, sebuah penglihatan murni.

\english
The night is full of stars. Lie down on the earth, disappear into the earth. We come from the earth, one day we will be going back to the earth to rest forever. At night sometimes, lying on the lawn, disappear into the earth. Look at the stars -- just look, a pure look.

\bahasa
Jangan mulai memikirkan nama bintang, nama rasi bintang. Lupakan semua yang engkau ketahui tentang bintang, sisihkan semua pengetahuanmu, lihat saja bintang-bintang tu. Dan tiba-tiba akan ada persekutuan; bintang-bintang akan mulai menuangkan cahaya ke dalam dirimu, dan engkau akan merasakan perluasan kesadaran. Tidak ada obat-obatan yang bisa melakukannya.

\english
Don't start thinking about the names of the stars, the names of the constellations. Forget all that you know about stars, put aside all your knowledge, just see the stars. And suddenly there will be a communion; the stars will start pouring their light into you, and you will feel an expanding of consciousness. No drug can do it.

\bahasa
Obat-obatan sangat tidaklah natural, buatan dan metode berbahaya untuk mengetahui sesuatu yang tersedia secara alami, tersedia dengan mudah, tersedia untuk hal yang berguna. Hanya menyaksikan bintang, engkau akan mulai merasa tinggi, engkau akan mulai membumbung tinggi.

\english
Drugs are very artificial, arbitrary and harmful methods to know something which is naturally available, which is easily available, beneficially available. Just watching the stars, you will start feeling high, you will start soaring high.

